Sensor Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems

### 2023-04-27 latex/intro.tex
getting there

\section{Introduction}\label{section:OED intro}
Experimental design is an important part of many scientific
investigations. When considering an inverse problem, one can often
specify sensor locations (e.g.\ in electricat impedance tomography
\cite{horesh2010impedance}), certain wavelengths (e.g.\ in MRI
\cite{horesh2008mri}), or location of sensors for waves reflecting
from the ground (e.g.\ searching for oil \cite{horesh2008borehole} or
using a radar). Whatever the allowed set of measurements is, one
should select the optimal observations to take, in order to increase
accuracy, reduce costs, or both.

A set of measurements is called a \emph{design}, and an \emph{optimal
design} is found by optimizing some \emph{design
criterion}. Surprisingly, optimal designs sometimes result in
measuremets that are very similar. For example consider measurements
taken by sensors that evaluate some function at $\x \in \Omega
\subseteq \R^d, d=1,2,3$. An optimal design may place sensors in very
close proximity to each other (Figure \ref{fig:clusterization
  illustration}). Following \cite{Ucinski05}, we refer to this
phenomenon as \emph{sensor clusterization}.

There is a consensus among researchers that sensor clusterization is
undesirable. Some authors acknowledge the problem of sensor
clusterization and suggest remedies \cite{fedorov1996, hooker2009,
  fedorov2012, Ucinski05, neitzel2019sparse}. Others bypass the
problem by allowing sensor locations to be chosen from a coarse finite
grid \cite{alexanderian2014sparsification,
  AlexanderianPetraStadlerEtAl16, alexanderian2018efficient,
  koval2020, attia2020}, thus eliminating the possibility of sensor
clusterization. Sadly, there is no free lunch, and choosing sensor
locations from a coarse finite grid requires solving a hard
combinatorial optimization problem. The current study is the first, to
my knowledge, that aims to gain a deep understanding of sensor
clusterization.

\subsection{D-Optimal Designs}
In this study we focus on the Bayesian \emph{D-optimality} criterion
\cite{Chaloner1995}: a D-optimal design maximizes the expected
Kullback-Leibler divergence \cite{CoverThomas91} between posterior and
prior. For a linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. For a linear
model with Gaussian noise, the Bayesian criterion is just
a regularized version of the frequentist criterion
\cite{Chaloner1995}.}.

Bayesian D-optimal designs for inverse problems on function spaces
were first studied by \cite{AlexanderianGloorGhattas14,
  AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, a D-optimal design maximizes the KL divergence between posterior
and prior measures. For a linear inverse problem with Gaussian prior
and Gaussian noise, a D-optimal design minimizes an expression
analogous to the determinant of the posterior covariance operator
\cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
below).

Considerable effort focused on computational aspects of finding
optimal designs in infinite-dimensional Bayesian inverse problems.
Attia \cite{attia2020} focused on QoI, Koval considered model
uncertainty and focused on reducing forward and adjoint solves, using
randomized linear algebra.\cite{koval2020}.

\subsection{Remedies to Sensor Clusterization}
Several remedies to sensor clusterization were suggeted through the
years, and we will now overview some. In \emph{clusterization-free
design} locations are forced to be far from each other
\cite{Ucinski05}. This is achieved in a frequentist and
finite-dimensional setting, by either imposing distance constraints
between observations, or by introducing correlated errors which
account for both observation error and model error
\cite{Ucinski05}. Merging close measurements is also possible
\cite{fedorov2012}. Sensor clusterization is mitigated in time-series
analysis for pharmacokinetic experiments by modeling auto-correlation
time in the noise \cite{hooker2009}, and this approach is similar in
spirit to imposing error correlations as in \cite{Ucinski05}.

Many authors bypass the sensor clusterization problem by restricting
measurements to a coarse finite grid \cite{koval2020,
  AlexanderianPetraStadlerEtAl16, alexanderian2021, attia2020}, but
this practice has two flaws: First, using a coarse grid requires
solving a hard combinatorial optimization problem. Second, restricting
sensors to a such a small subset of the computatoinal domain will
necessarily give rise to sub-optimal designs. Several remedies were
offered for the first problem: Stadler suggested ell0
sparsification. Others suggested other stuff. None were offered for
the second.

\subsection{Contribution}
Any of the above mentioned approaches might serve as a remedy and push
sensors away from each other. Yet, none offers any insight as to why
clusterization occurs. Also, as better models are employed, model
error is decreased and the clusterization phenomenon will eventually
reappear. While these approaches are practical and help practitioners
avoid sensor clusterization, they do not provide insight as to why
sensors are clustering.

The current study aims to fill the gap of understanding sensor
clusterization in infinte dimensional Bayesian linear inverse
problems. We propose and study a relaxed and analytically tractable
model for D-optimal designs. Under this model, D-optimal designs are
solutions of a constrained optimization problem, formulated using
Lagrange multipliers. The Lagrange multipliers formulation allows us
to rigorously show how model error mitigates clusterization (Section
\ref{section:non vanishing}), and why designs that exhibit
clusterization are indeed optimal when no model error is considered.

  Let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on $\hilp$,
    where $\prcov:\hilp \to \hilp$ is the prior covariance operator.
  \item $\obs: \hilo \to \mathbb{R}^m$ observation operator, where $m
    \in \mathbb{N}$ is the number of observations taken. 
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance,
    $\data = \obs \fwd \param + \eps$, where $\eps \in \mathbb{R}^m$
    is iid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality utility function
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
    in decreasing order of magnitude.
  %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \item $\obs$ a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs).
    \end{equation*}
  \end{itemize}

One byproduct of the proof of Theorem \ref{thm:char} is a novel lemma
in linear algebra: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

\subsection{An Example of Clusterization}\label{subsec:example}
\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      title={Posterior Pointwise Standard Deviations and D-Optimal Sensor Locations},  
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      %% \addplot [thin, black, mark=none] table{stdv-heat-sens1-var1.txt};
      %% \addlegendentry{1 sensors};

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Sensor clusterization for inferring the initial condition
    of the 1D heat equation. Posterior pointwise standard deviations
    (lines) are plotted over the domain $[0, \pi]$, for varying
    numbers of sensors. Sensor locations (circles) were chosen to
    optimize the D-optimality criterion. Sensor clusterization occurs
    for six sensors. Only four observation locations are used --- two
    pairs of sensors are so close together they are
    indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

For the purpose of illustration, consider a toy inverse problem ---
inferring the initial condition of the 1D heat equation in $[0,\pi]$
with a homogeneous Dirichlet boundary condition. Sensor clusterization
is clearly evident, see Figure \ref{fig:clusterization}.

@@

The goal is to infer the initial condition $u_0$. Model the initial
condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. Fix
some final time $T$ and denote by $\fwd$ the forward operator, so that
$u(\cdot, T) = \fwd u_0$. Let $\obs$ the observation operator, so that
$u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. Assume iid centered Gaussian
observation error, so observations are given by $u(\x_j,T) +
\eps(\x_j)$ with $\eps(\x_j) \sim \normal(0, \sigma^2)$ iid, $\sigma^2
> 0$. The posterior covariance is known and depends only on $\prcov,
\fwd, \obs$ and $\sigma^2$ \cite{Stuart10} (see Section
\ref{section:prelim} and \eqref{eq:postcov} below). Finally, choose
$\x_j,j=1,\dots,m$ to optimize the D-optimality criterion
\cite{AlexanderianGloorGhattas14}.

>>>>

\section{Introduction}\label{section:OED intro}
Experimental design is an important part of many scientific
investigations. When considering an inverse problem, one can often
specify sensor locations (e.g.\ in electricat impedance tomography
\cite{horesh2010impedance}), certain wavelengths (e.g.\ in MRI
\cite{horesh2008mri}), or location of sensors for waves reflecting
from the ground (e.g.\ searching for oil
\cite{horesh2008borehole}). Whatever the allowed set of measurements
is, one should select the optimal measurements to take, in order to
increase accuracy, reduce costs, or both. A set of measurements is
called a \emph{design}, and an \emph{optimal design} is found by
optimizing some \emph{design criterion}. The most popular desgin
criteria are A- and D-optimality

Surprisingly, optimal designs sometimes result in measuremets that are
very similar \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}. As an example, consider the inverse problem of
inferring the initial condition of the 1D heat equation (details in
Section \ref{subsec:example}). In Figure \ref{fig:clusterization
  illustration} we see that an optimal design results in two sets of
sensors that are almost indistinguishable from each other. Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
clusterization}.

\begin{figure}
  \begin{tikzpicture}[thick, scale=1.3, every node/.style={scale=0.99}]
    \begin{axis}
      [
      title={Posterior Pointwise Standard Deviations and D-Optimal Sensor Locations},  
      xmin = 0,
      xmax = 3.14,
      xlabel = {$x \in \Omega$},
      ylabel = posterior std,
      ymin   = 0,
      %compat = 1.3,
      % ymax   = 130,
      % ytick = \empty,
      legend cell align=left,
      % legend style={at={(0.45,0.2)}}
      legend pos= outer north east 
      ]
      % \draw[black!30!white, thin] (50,0) -- (50,130);
      % 
      %% \addplot [thin, black, mark=none] table{stdv-heat-sens1-var1.txt};
      %% \addlegendentry{1 sensors};

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Sensor clusterization for inferring the initial condition
    of the 1D heat equation. Posterior pointwise standard deviations
    (lines) are plotted over the computatoinal domain $\Omega = [0,
      \pi]$, for varying numbers of sensors. Sensor locations
    (circles) were chosen to optimize the D-optimality
    criterion. Sensor clusterization occurs for six sensors: Only four
    observation locations are visible, since two pairs of sensors are
    so close together they are indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

There is a consensus among researchers that sensor clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse} and several remedies to sensor clusterization
were suggeted through the years. In a \emph{clusterization-free
design}, sensor locations are forced to be far from each other
\cite{Ucinski05}. In a frequentist and finite-dimensional setting,
senesor clusterization is avoided by either imposing distance
constraints between measurements, or by introducing correlated errors
which account for both observation error and model error
\cite{Ucinski05}. Merging close measurements is also possible
\cite{fedorov2012}, but this practice ignores sensor clusterization
altogether. In time-series analysis for pharmacokinetic experiments,
sensor clusterization is mitigated by modeling auto-correlation time
in the noise \cite{hooker2009}, and this approach is similar in spirit
to assuming errors are correlatied as in \cite{Ucinski05}.

%% Sometimes, imposing a correlation structure on observation noise is
%% reasonable indeed \cite{koval2020, Alexanderian2021}. For example,
%% Koval \cite{koval2020} imposes a Gaussian prior on ocean bathytermy
%% when inferring BLA. However, in the (un)fortunate event that ocean
%% bathytermy is known, there is no use to modeling it as an unknown
%% and sensor clusterization will rear its ugly head.

Many scientists bypass the sensor clusterization problem by
restricting measurements to a coarse grid of the computational domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$ \cite{koval2020,
  alexanderian2021, attia2020, alexanderian2014sparsification,
  AlexanderianPetraStadlerEtAl16, alexanderian2018efficient}, but this
practice requires solvilg a hard combinatorial optimization problem
for choosing optimal sensor locations from a finite set. A solution
typically involves relaxing the problem to that of choosing optimal
measurement ``weight'' in $[0,1]$ for each potential sensor
location. Some authors then add a sparsifying $\ell_1$ penalty term to
the design criterion, which is subsequently thresholded
\cite{horesh2008borehole}, thus resulting in a sparse binary
solution. Others successively relax the $\ell_1$ penalty to a $\ell_0$
penalty via a continuation method
\cite{AlexanderianPetraStadlerEtAl16,
  alexanderian2014sparsification}. A clever alternative is to cast the
problem of finding optimal sensor weights as a stochastic optimization
problem, which in turn is solved by stochastic gradient descent
\cite{attia2022stochastic}. While all of the above mentioned methods
are valid solutions to the prohibitive computational cost of finding
an binary optimal design restricted to a coarse grid, none treats the
elephant in the room: restricting sensors to a coarse grid in $\Omega
\subseteq \mathbb{R}^d, d=1,2,3$ necessarily gives rise to a
sub-optimal design.

Avoiding sensor clusterization is pragmatical: we feel sensor
clusterization is bad (but do not know why), and try to avoid
it. However, all of the abovementioned techniques that avoid sensor
clusterization do so by somehow restricting sensor locations, thus
fundamentaly altering the optimal design problem. To the best of my
knowledge, no study tried to answer the following seemingly trivial
and basic questions: why do sensor clusterization occurs? Is sensor
clusterization really bad? Can we find optimal designs BLABLA?

%% However, as far as I am aware, no study concerned the fact that
%% restricting sensors to a coarse grid is, by definition,
%% sub-optimal.

%% Attia \cite{attia2020} focused on QoI, Koval considered model
%% uncertainty and focused on reducing forward and adjoint solves, using
%%  randomized linear algebra \cite{koval2020}.

%% Considerable efforts focused on various computational aspects of
%% finding optimal designs in infinite-dimensional Bayesian inverse
%% problems \cite{attia2020, koval2020,
%% AlexanderianPetraStadlerEtAl16, attia2023pyoed, koval2021optimal,
%% attia2020stochastic}, the subject of sensor clusterization was less
%% investigated, and

%% Second, restricting sensors to a such a small subset of the computatoinal
%% domain $\Omega \subseteq \mathbb{R}^d, d=1,2,3$ necessarily gives rise
%% to sub-optimal designs.

\subsection{Contribution}
In this study, we seeks to understand sensor clusterization. We focus
our attention on the Bayesian \emph{D-optimality} criterion: a
D-optimal design maximizes the expected Kullback-Leibler divergence
\cite{CoverThomas91} between posterior and prior
\cite{Chaloner1995}. We study Bayesian D-optimal designs in linear
inverse problems that are (potentially) of infinte dimensions. We
propose and study a relaxed and analytically tractable model for such
D-optimal designs. Under this model, D-optimal designs are solutions
of a constrained optimization problem, formulated using Lagrange
multipliers. The Lagrange multipliers formulation allows us to
rigorously show why model error mitigates clusterization (Section
\ref{section:non vanishing}), like many authors note, and why designs
that exhibit clusterization are indeed optimal when no model error is
considered (Section \ref{section:vanishing}).

  Let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on $\hilp$,
    where $\prcov:\hilp \to \hilp$ is the prior covariance operator.
  \item $\obs: \hilo \to \mathbb{R}^m$ observation operator, where $m
    \in \mathbb{N}$ is the number of observations taken. 
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance,
    $\data = \obs \fwd \param + \eps$, where $\eps \in \mathbb{R}^m$
    is iid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
    in decreasing order of magnitude.
  %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \item $\obs$ a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs).
    \end{equation*}
  \end{itemize}

Another byproduct of the proof of Theorem \ref{thm:char} is a novel
linear algebra lemma: We show how to decompose a symmetric positive
definite matrix $M$ as $M = AA^t$, where $A$ has unit norm columns
(Lemma \ref{lemma:lax}). We also generalize two lemmas from linear
algebra to infinite-dimensional settings: In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces, and in Lemma
\ref{lemma:lax} we generalize a lemma \cite{Lax07} for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function.

\subsection{An Example of Clusterization}\label{subsec:example}
For the purpose of illustration, consider a toy inverse problem ---
inferring the initial condition of the 1D heat equation in $\Omega :=
[0,\pi]$ with a homogeneous Dirichlet boundary condition. Sensor
clusterization is clearly evident, see Figure \ref{fig:clusterization
  illustration}.

@@

The goal is to infer the initial condition $u_0$. Model the initial
condition as $u_0 \sim \normal(0,\prcov)$, for $\prcov =
(-\Delta)^{-1}$ with a homogeneous Dirichlet boundary condition. Fix
some final time $T$ and denote by $\fwd$ the forward operator, so that
$u(\cdot, T) = \fwd u_0$. Let $\obs$ the observation operator, so that
$u(\x_j,T) = (\obs u)_j, j=1,\dots,m$. Assume iid centered Gaussian
observation error, so observations are given by $u(\x_j,T) +
\eps(\x_j)$ with $\eps(\x_j) \sim \normal(0, \sigma^2)$ iid, $\sigma^2
> 0$. The posterior covariance is known and depends only on $\prcov,
\fwd, \obs$ and $\sigma^2$ \cite{Stuart10} (see Section
\ref{section:prelim} and \eqref{eq:postcov} below). Finally, choose
$\x_j,j=1,\dots,m$ to optimize the D-optimality criterion, see Theorem
\ref{thm:d optimality} \cite{AlexanderianGloorGhattas14}.

### 2023-04-27 latex/prelims.tex
getting there

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, $\fwd \prcov \fwd^*$ is assumed invertible
--- an assumption which will be used below. Let $m$ the number of
observations taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly
separated \cite{cvetkovic2023choosing}.} $\obs \in ( \hilo^* )^m$, so that:

@@

D-optimal designs in infinite-dimensional Hilbert spaces were
investigated in \cite{AlexanderianGloorGhattas14}. Their main result
is summarized, in the notation presented above, in the following
theorem\footnote{In \cite{AlexanderianGloorGhattas14}\footnote{A
similar theorem appears in \cite[Theorem
  1]{alexanderian2018efficient}.}, results are stated for $\Sigma=I$
(implied by $\modcov = 0,\sigma^2=1$), but the authors also mention
these results hold for more general covariance matrices (p. 681).}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

@@

\subsection{Notation Summary}\label{subsec:notation}
  Our notation is summarized below. We let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on
    $\hilp$, with prior covariance operator $\prcov:\hilp \to \hilp$.
  \item $\obs: \hilo \to \mathbb{R}^m$ observation operator, where $m
    \in \mathbb{N}$ is the number of observations taken. 
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance.
  \item $\modcov$ model error covariance operator.
    %% $\data = \obs \fwd \param + \eps$, where $\eps \in
    %% \mathbb{R}^m$ isiid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality utility function
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
        \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  %% \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
  %%   in decreasing order of magnitude.
  %% %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  %% \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \end{itemize}

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectfully). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator from section
\ref{subsec:example} is a prime example). Take a Gaussian prior
$\param \sim \pr = \normal(\prmean ,\prcov)$ with some appropriate
covariance operator $\prcov$ \cite{Stuart10} on $\hilp$. Note that
$\fwd \prcov \fwd^*$ is the prior covariance in $\hilo$
\cite{Stuart10}. As such, $\fwd \prcov \fwd^*$ is assumed invertible
--- an assumption which will be used below. Let $m$ the number of
observations taken. Observations are taken via the \emph{observation
operator}\footnote{It is common for the observation and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly separated
similarly to \cite{attia2022stochastic, cvetkovic2023choosing}.} $\obs
\in ( \hilo^* )^m$, so that:

@@

For a Bayesian linear model in finite dimensions, with Gaussian prior
and Gaussian noise, a D-optimal design minimizes the determinant of
the posterior covariance matrix\footnote{The frequentist version of
the D-optimality criterion aims to minimize the determinant of the
Fisher information matrix \cite[page 16]{Ucinski05}. Thus, the
Bayesian criterion is just a regularized version of the frequentist
criterion \cite{Chaloner1995}.}. Bayesian D-optimal designs for
inverse problems on function spaces were first studied by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}, and just like in finite dimensions,
a D-optimal design maximizes the Kullback-Leibler divergence between
posterior and prior measures.
%% For a linear inverse problem with Gaussian prior and Gaussian
%% noise, a D-optimal design minimizes an expression analogous to the
%% determinant of the posterior covariance operator
%% \cite{AlexanderianGloorGhattas14} (also \cref{thm:d optimality}
%% below).
D-optimal designs in infinite-dimensional
Hilbert spaces were investigated in
\cite{AlexanderianGloorGhattas14}. Their main result is summarized, in
the notation presented above, in the following theorem\footnote{In
\cite{AlexanderianGloorGhattas14, alexanderian2018efficient}, results
are stated for $\Sigma=I$ (implied by $\modcov = 0,\sigma^2=1$), but
the authors also mention these results hold for more general
covariance matrices \cite[p. 681]{AlexanderianGloorGhattas14}.}:
\begin{theorem}[Slightly modified Theorem 1 from \cite{AlexanderianGloorGhattas14}]\label{thm:d optimality}
  Let $\pr = \normal(\prmean,\prcov)$ be a Gaussian prior on $\hilp$
  and let $\post = \normal(\postmean,\postcov)$ the posterior measure
  on $\hilp$ for the Bayesian linear inverse problem $\data = \obs
  \fwd\param + \obs \eps' + \eps$ discussed above. Then
  \begin{align}\label{eq:objective}
    \begin{split}
      \tar( \obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
      % 
      % 
      % 
      &= \frac12 \log \det 
      ( I + \prcov^{1/2}  \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}).
    \end{split}
  \end{align}
\end{theorem}
\begin{definition}\label{def:d optimality}
  $\obs^{\star}$ is said to be \emph{D-optimal} if $\obs^{\star} =
  \argmax_{\obs} \tar(\obs)$, where entries of $\obs \in (\hilo^*)^m$ are constrained to
  some allowed set of measurement vectors in $\hilo^*$.
\end{definition}

@@

\subsection{Notation Summary}\label{subsec:notation}
  Our notation is summarized below. We let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on
    $\hilp$, with prior covariance operator $\prcov:\hilp \to \hilp$.
  \item $\obs: \hilo \to \mathbb{R}^m$ observation operator, where $m
    \in \mathbb{N}$ is the number of observations taken. 
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance.
  \item $\modcov$ model error covariance operator.
    %% $\data = \obs \fwd \param + \eps$, where $\eps \in
    %% \mathbb{R}^m$ isiid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\Sigma(\obs) = \obs \modcov \obs^* + \sigma^2I$. 
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \prcov^{1/2} \fwd ^* \obs^* \Sigma(\obs)^{-1} \obs
        \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  %% \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
  %%   in decreasing order of magnitude.
  %% %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  %% \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \end{itemize}
